Dewlap color variation in \textit{Anolis sagrei} is maintained among habitats within islands of the West Indies

### 2021-09-06 ms/main_title.tex
stuff

\title{Dewlap color variation in \textit{Anolis sagrei} is maintained among habitats within islands of the West Indies}
\author{
    \textsc{Rapha\"{e}l Scherrer$^{1,3}$ \thanks{Corresponding author: r.scherrer@rug.nl}, Colin M. Donihue$^{1,4}$, }\\
	\textsc{R. Graham Reynolds$^2$, Jonathan B. Losos$^{1,4}$ and Anthony J. Geneva$^{1,5}$} \\[1ex]
	%\thanks{A thank you} \\[1ex] % Your name
	\normalsize $^1$ Department of Organismic and Evolutionary Biology and Museum of Comparative Zoology \\ \normalsize Harvard University, Cambridge, MA, USA \\
	\normalsize $^2$ Department of Biology, University of North Carolina Asheville, Asheville, NC, USA\\ 
	\normalsize $^3$ Current address: Groningen Institute for Evolutionary Life Sciences,\\
	\normalsize University of Groningen, Groningen, The Netherlands\\
	\normalsize $^4$ Current address: Department of Biology, Washington University, St. Louis, MO, USA\\
	\normalsize $^5$ Current address: Department of Biology, Center for Computational and Integrative Biology,\\ \normalsize Rutgers University--Camden, Camden, NJ, USA
}
\date{} % Leave empty to omit a date

@@

\begin{abstract}
    \input{sections/abstract.tex}
\end{abstract}

\textbf{Keywords} --- reflectance, adaptation, sexual signal, machine learning, polymorphism

\section*{Introduction}

\input{sections/introduction.tex}

\pagebreak

\section*{Methods}

\input{sections/methods2.tex}

\pagebreak

\section*{Results}

\input{sections/results2.tex}

\pagebreak

\section*{Discussion}

\input{sections/discussion2.tex}

\pagebreak

\subsection*{Acknowledgements}

\input{sections/acknowledgments.tex}

\pagebreak

\section*{Figures}

\input{sections/figures.tex}

\pagebreak

\section*{Tables}

\input{sections/tables.tex}

\pagebreak

\bibliographystyle{apalike}
\bibliography{library.bib}

\beginsupplement

\pagebreak

\section*{Appendix}

\input{sections/appendix}

\pagebreak

\section*{Supplementary Figures}

\input{sections/suppfigures}

\pagebreak

\section*{Supplementary Tables}

\input{sections/supptables.tex}

>>>>

\title{Dewlap color variation in \textit{Anolis sagrei} is maintained among habitats within islands of the West Indies}
\author{
	\textsc{Rapha\"{e}l Scherrer}$^{1,3}$ \thanks{Corresponding author} (raphael.scherrer@evobio.eu, ORCID 0000-0002-1447-7630),\\
    \textsc{Colin M. Donihue}$^{1,4}$ (colindonihue@gmail.com, 0000-0003-1096-8536),\\
	\textsc{R. Graham Reynolds}$^2$ (greynold@unca.edu, 0000-0002-0056-8304),\\ 
    \textsc{Jonathan B. Losos}$^{1,4}$ (losos@wustl.edu, 0000-0003-4712-8258) and\\
	\textsc{Anthony J. Geneva}$^{1,5}$ (anthony.geneva@gmail.com, 0000-0001-8253-6527)\\[1ex]
	\normalsize $^1$ Department of Organismic and Evolutionary Biology and Museum of Comparative Zoology \\ \normalsize Harvard University, Cambridge, MA, USA \\
	\normalsize $^2$ Department of Biology, University of North Carolina Asheville, Asheville, NC, USA\\ 
	\normalsize $^3$ Current address: Groningen Institute for Evolutionary Life Sciences,\\
	\normalsize University of Groningen, Groningen, The Netherlands\\
	\normalsize $^4$ Current address: Department of Biology, Washington University, St. Louis, MO, USA\\
	\normalsize $^5$ Current address: Department of Biology, Center for Computational and Integrative Biology,\\ \normalsize Rutgers University--Camden, Camden, NJ, USA
}
\date{} % Leave empty to omit a date

@@

	\textbf{Running title} --- Dewlap coloration in \textit{Anolis sagrei}.

### 2021-09-06 ms/sections/discussion2.tex
stuff

Detectable differences in dewlap color between populations are surprising, as habitats were often in close geographical proximity to each other (as close as a few hundred meters on Bimini and most of the time within ten kilometers). Indeed, given that (1) the populations were continuously distributed between the habitats, (2) populations from different habitats were not monophyletic with respect to mitochondrial haplotypes (\citealt{vandeSchoot2016} unpublished thesis), and (3) \textit{A. sagrei} has been shown to be a highly mobile species \citep{Kamath2018}, we would have expected more homogeneous distributions of color phenotypes within islands due to extensive gene flow, with fewer differences between populations, especially those in close proximity.\\

Several scenarios could account for these findings. One explanation is an adaptive one: populations living in different habitats could be phenotypically adapted to their local environmental conditions. Given that the brightly colored dewlap of \textit{A. sagrei} is used as a communication signal, its color may be a target for selection if the transmission or perception of the signal differs from one habitat to another, for example because of differences in ambient light, according to the sensory-drive hypothesis \citep{Endler1988, Endler1992, Endler1998}. The sensory-drive hypothesis has been tested multiple times for dewlap coloration in \textit{Anolis} lizards, with mixed results. Some authors found support for it \citep{Leal2002, Leal2004}, while others did find differences in dewlap coloration between habitats, but those were inconsistent with the sensory-drive hypothesis \citep{Fleishman2009, Ng2012}.\\

If our results were an example of sensory drive, we would have expected to see consistent differences between populations from different habitats across islands, given the apparent environmental consistency each of the three habitat types across the islands we sampled. In particular, we would have expected divergence in line with increased detectability given local light conditions, such as the high contrasts with background vegetation found in the UV range in \citet{Leal2002} and \citet{Leal2004}. We might also have expected mangrove and beach scrub lizards, both inhabiting areas with high light penetration, to harbor more similar dewlaps, and to differ significantly form lizards from the coppice habitat, where irradiance is low. Instead, we found inconsistencies in the way dewlap color differed between habitats across islands. While short-wavelengths (UV reflectance) were often involved in color differences, they were not involved on all islands where significant differences were detected. On some islands, other or additional variables differed, such as brightness, red reflectance or the reflectance at the ends of the spectrum visible to \textit{Anolis} lizards (UV and red, \citealt{Lazareva2012}) relative to intermediate wavelengths (blue-to-yellow). Similar portions of the spectrum were sometimes involved in opposite directions on different islands, such as on Abaco and Cayman Brac, where mangrove lizards had a higher UV-reflectance than beach scrub lizards on the former, but a lower UV-reflectance on the latter. Overall, the observed heterogeneity of divergence patterns across islands provides no support to a sensory-drive explanation.\\

@@

We found no evidence for a role of genetic drift in explaining the observed patterns either. First, \textit{A. sagrei} was distributed across the islands continuously, usually at relatively high population densities, and not in small and isolated populations where drift might be expected to have a strong effect. Second, we found no evidence of isolation-by-distance except on Abaco, which was sampled at the largest geographical scale, with sites nearly a hundred kilometers apart from each other. Hence, while isolation-by-distance may explain long-range differences on this island, most of the differences among habitats across islands, which were often in close proximity (less than 10km), are unlikely to be explained by genetic drift.\\

In this study, we found larger differences among than within islands, a pattern already reported and linked to climatic conditions \citep{Vanhooydonck2009} and to densities of predators and of anole congeners \citep{Baeckens2018}. Differences among habitats within islands, however, are still difficult to account for. Remaining hypotheses may include, for example, runaway sexual selection arbitrarily operating in different directions across islands \citep{Andersson1994}, but no evidence so far suggests that dewlap is a target of mate choice in anoles \citep{Tokarz2002, Tokarz2005, Nicholson2007}. Another hypothesis is that the different genetic constitutions of different islands, perhaps resulting from founder effects (the islands have been colonized independently from Cuba, \citealt{vandeSchoot2016} unpubl.; \citealt{Driessens2017, Reynolds2020}), may have predisposed populations to adapt differently to similarly selective circumstances. Either way, new data would be needed to test these hypotheses.\\

Altogether, our results show that dewlap color of \textit{A. sagrei} commonly varies between habitat types, even in close geographical proximity, within islands of the West Indies. Besides, coloration differs in different ways across similar habitats from one island to another. We discussed several non-mutually exclusive mechanisms that could explain these observations, but more data are needed to investigate these scenarios. Nevertheless, heterogeneous patterns of divergence across islands do not support an adaptive sensory-drive scenario, and we propose that within-island dewlap color variation may be underlain by a more subtle mosaic of factors.

>>>>

Detectable differences in dewlap color between populations are surprising, as habitats were often in close geographical proximity to each other (as close as a few hundred meters on Bimini and most of the time within ten kilometers). Indeed, given that (1) the populations were continuously distributed between the habitats, (2) populations from different habitats were not monophyletic with respect to mitochondrial haplotypes (\citealt{vandeSchoot2016} unpublished thesis), and (3) \textit{A. sagrei} is highly mobile \citep{Kamath2018}, we would have expected more homogeneous distributions of color phenotypes within islands due to gene flow, with fewer differences between populations, especially those in close proximity.\\

Several scenarios could account for these findings. One explanation is an adaptive one: populations living in different habitats could be phenotypically adapted to their local environmental conditions. Given that the brightly colored dewlap of \textit{A. sagrei} is used as a communication signal, its color may be a target for selection if the transmission or perception of the signal differs from one habitat to another, for example because of differences in ambient light, according to the sensory-drive hypothesis \citep{Endler1988, Endler1992, Endler1998}. The sensory-drive hypothesis has been tested multiple times for dewlap coloration in \textit{Anolis} lizards, with mixed results. Some authors found support for it \citep{Leal2002, Leal2004}, while others found differences in dewlap coloration between habitats inconsistent with the sensory-drive hypothesis \citep{Fleishman2009, Ng2012}.\\

If our results were an example of sensory drive, we would have expected to see consistent differences between populations from different habitats across islands, given the apparent environmental consistency each of the three habitat types across the islands we sampled. In particular, we would have expected divergence in line with increased detectability given local light conditions, such as the high contrasts with background vegetation found in the UV range in \citet{Leal2002} and \citet{Leal2004}. We might also have expected mangrove and beach scrub lizards, both inhabiting areas with high light penetration, to have more similar dewlaps, and to differ significantly from lizards from the coppice habitat, where irradiance is low. Instead, we found inconsistencies in the way dewlap color differed between habitats across islands. While short-wavelengths (UV reflectance) were often involved in color differences, they were not involved on all islands where significant differences were detected. On some islands, other or additional variables differed, such as brightness, red reflectance or the reflectance at the ends of the spectrum visible to \textit{Anolis} lizards (UV and red, \citealt{Lazareva2012}) relative to intermediate wavelengths (blue-to-yellow). Similar portions of the spectrum were sometimes involved in opposite directions on different islands, such as on Abaco and Cayman Brac, where mangrove lizards had a higher UV-reflectance than beach scrub lizards on the former, but a lower UV-reflectance on the latter. Overall, the observed heterogeneity of divergence patterns across islands provides no support to a sensory-drive explanation.\\

@@

We found no evidence for a role of genetic drift in explaining the observed patterns either. First, \textit{A. sagrei} was distributed across the islands continuously, usually at relatively high population densities, rather than in small and isolated populations where drift might be expected to have a strong effect. Second, we found no evidence of isolation-by-distance except on Abaco, which was sampled at the largest geographical scale, with sites nearly a hundred kilometers apart from each other. Hence, while isolation-by-distance may explain long-range differences on this island, most of the differences among habitats across the rest of the sampling region are unlikely to be explained by genetic drift, as habitats were often in close proximity (less than 10km).\\

In this study, we found larger differences among than within islands, a pattern already reported and linked to climatic conditions \citep{Vanhooydonck2009} and to densities of predators and of anole congeners \citep{Baeckens2018}. Differences among habitats within islands, however, are still difficult to account for. Remaining hypotheses may include, for example, runaway sexual selection arbitrarily operating in different directions across islands \citep{Andersson1994}, but no evidence so far suggests that dewlap is a target of mate choice in anoles \citep{Tokarz2002, Tokarz2005, Nicholson2007}. Another hypothesis is that the different genetic constitutions of different islands, perhaps resulting from founder effects (the islands have been colonized independently, \citealt{vandeSchoot2016} unpubl.; \citealt{Driessens2017, Reynolds2020}), may have predisposed populations to adapt differently to similarly selective circumstances. Either way, new data would be needed to test these hypotheses.\\

Altogether, our results show that dewlap color of \textit{A. sagrei} commonly varies between habitat types, even in close geographical proximity, within islands of the West Indies. However, coloration differs in different ways across similar habitats from one island to another. We discussed several non-mutually exclusive mechanisms that could explain these observations. Nevertheless, heterogeneous patterns of divergence across islands do not support an adaptive sensory-drive scenario, and we propose that within-island dewlap color variation may be underlain by a more subtle mosaic of factors.

### 2021-09-06 ms/sections/introduction.tex
stuff

The species \textit{Anolis sagrei} is widespread across islands of the West Indies \citep{Reynolds2020}. It has been the subject of numerous studies concerning local adaptation \citep{Losos1994, Losos1997a, Losos2001, Kolbe2012}, biological invasion \citep{Kolbe2008}, and sexual selection \citep{Tokarz2002, Tokarz2005, Tokarz2006, Driessens2014, Steffen2014, Driessens2015} among many other topics. Between-island variation in the mainly orange-red color of its dewlap was shown to be better explained by climatic variables \citep{Driessens2017} than by proxies for biotic factors such as sexual selection or predation pressure \citep{Vanhooydonck2009, Baeckens2018}. How intra-island differences in habitat may contribute to the diversity of dewlap coloration, however, remains unexplored, and may reveal new insights into the scale of local differentiation despite gene flow.\\

@@

Our sampling design included sites in close proximity; the median distance between two sites within an island was $11.2$km. Combining reflectance spectrometry and supervised machine learning, we tested for divergence in dewlap phenotype between habitats within islands and between islands across part of the range of \textit{A. sagrei}. We predicted that if light conditions in the environment indeed drive color evolution, dewlaps should be most similar between beach scrub and mangrove forest, which both have high levels of light irradiance, compared to the darker, closed-canopy coppice forest. Similar, if detectability is maximized given the local conditions, we expected darker and more contrasting dewlaps in high irradiance habitats. Finally, if habitat characteristics are strong determinants of dewlap color variation, similar patterns should be observed across multiple islands \citep{Harvey1991, Losos2011}.

>>>>

\textit{Anolis sagrei} is widespread across islands of the West Indies \citep{Reynolds2020}. It has been the subject of numerous studies concerning local adaptation \citep{Losos1994, Losos1997a, Losos2001, Kolbe2012}, biological invasion \citep{Kolbe2008}, and sexual selection \citep{Tokarz2002, Tokarz2005, Tokarz2006, Driessens2014, Steffen2014, Driessens2015} among many other topics. Between-island variation in the mainly orange-red color of its dewlap was shown to be better explained by climatic variables \citep{Driessens2017} than by proxies for biotic factors such as sexual selection or predation pressure \citep{Vanhooydonck2009, Baeckens2018}. How intra-island differences in habitat may contribute to the diversity of dewlap coloration, however, remains unexplored, and may reveal new insights into the scale of local differentiation despite gene flow.\\

@@

Our sampling design included sites in close proximity; the median distance between two sites within an island was $11.2$km. Combining reflectance spectrometry and supervised machine learning, we tested for divergence in dewlap phenotype between habitats within islands and between islands across part of the range of \textit{A. sagrei}. We predicted that if light conditions in the environment indeed drive color evolution, dewlaps should be most similar between beach scrub and mangrove forest, which both have high levels of light irradiance, compared to the darker, closed-canopy coppice forest. If detectability is maximized given the local conditions, we expected darker and more contrasting dewlaps in high irradiance habitats. Finally, if habitat characteristics are strong determinants of dewlap color variation, similar patterns should be observed across multiple islands \citep{Harvey1991, Losos2011}.

### 2021-09-06 ms/sections/results2.tex
stuff

We subsequently tested for differences in dewlap coloration between habitat-populations within each island, using within-island principal component scores (to maximize the variation captured for each island, see Methods). Our within-island random forest classification analyses revealed detectable differences in dewlap coloration on eight out of the nine islands in our sample: Abaco, Bimini, Cayman Brac, Eleuthera, Little Cayman, Long Island, North Andros and South Andros. The accuracy of random forest classification exceeded random expectation more often than expected by chance for all these islands (Table \ref{tab:randomforests}). Accuracy was as high as 73\% for Cayman Brac. We obtained similar results using other machine learning approaches such as support vector machines (Table \ref{tab:ksvms}) and linear discriminant analysis (Table \ref{tab:ldas}), except that these methods did not detect significant differences on Eleuthera and North Andros. We describe in details the specific differences detected on each island in the Appendix, and instead focus here on the general patterns emerging from our data.\\

>>>>

We subsequently tested for differences in dewlap coloration between habitat-populations within each island, using within-island principal component scores (to maximize the variation captured for each island, see Methods). Our within-island random forest classification analyses revealed detectable differences in dewlap coloration on eight out of the nine islands in our sample: Abaco, Bimini, Cayman Brac, Eleuthera, Little Cayman, Long Island, North Andros and South Andros. The accuracy of random forest classification exceeded random expectation more often than expected by chance for all these islands (Table \ref{tab:randomforests}). Accuracy was as high as 73\% for Cayman Brac. We obtained similar results using other machine learning approaches such as support vector machines (Table \ref{tab:ksvms}) and linear discriminant analysis (Table \ref{tab:ldas}), except that these methods did not detect significant differences on Eleuthera and North Andros. We describe in details the specific differences detected on each island in the Appendix, and focus here on the general patterns emerging from our data.\\
